Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-11-29 manuscript/manuscript_file.Rmd
Most recent updates to manuscript

## NOTES

-   incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
-   Citation "epa_data" isn't populating citation 11 in bibliography. The bibTex entry is classified as a "website". Roy, can you attempt to make this citation play nice with nature_geosciences.csl?
-   Letters vs article format at Nature Geosciences
    + Letters no longer supported.  Of all the options, Article is best suited to us.
    + 2200 – 3000 words (excluding abstract, Methods, references and figure legends)
    + Pseudo-abstract, 200 words, no citations
    + Subheadings are OK
    + Instructions to authors list separate Results and Discussion, but article in current issue do not follow this format.
    + Figures: article in current issue have 2-5 figures, mostly multi-panel.  Instructions specify 4-6 figures + tables.
-   Add quantitative NO3 values to Fig.3B and Fig. 5B legend.
-   Lester Luan and Jana Compton for internal tech reviews.

@@

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020] have called the accuracy of emission estimates into question [@Lauerwald_etal_2019]. Here we measured N~2~O concentrations and calculated emission rates for `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies distributed across the nine major ecoregions of the conterminous U.S. (CONUS). We use a Bayesian hierarchical approach to model the data and predicted N~2~O in`r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull()` CONUS waterbodies.  We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r propN2oSinkMean` % (95% credible interval:`r propN2oSinkLower` - `r propN2oSinkUpper`% ) of waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, and none of the ecoregions are estimated to function as net N~2~O sinks. The model predicts that waterbodies with high nitrate (NO~3~) tend to be sources of N~2~O whereas those with low NO~3~ are generally sinks. Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^. The mean estimate is a net emission of `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^, which is small relative to other known sources of N2O in the CONUS (e.g., wastewater treatment, manure management) and equivalent to only `r (national_stats %>% select(UCL_fn2o) %>% pull() / 22767) * 100 %>% round(., 1)`% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates . Emission estimates were most uncertain for large waterbodies, suggesting that efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nation’s largest lakes and reservoirs.  Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Nitrous oxide dynamics have been investigated in streams and rivers spanning a broad range of environemtal conditions, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that were at least 1 hectare in area, at least 1 meter deep, had at least 0.1 hectare of open water, and had a minimum residence time of one week[@EPAdesign] **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. These data, along with the survey design information, were used to develop a Bayesian hierarchical model to predict N~2~O concentrations for all 465,897 CONUS waterbodies in the target population. At the population level, the majority of CONUS waterbodies (mean = `r propN2oSinkMean`%, 95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) were predicted to be a net sink for N~2~O. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), although the 95% credible intervals overlap across all ecoregions **[(Table 1)]**.

@@

##### NO~3~

Dissolved N~2~O concentration increased with NO~3~ concentration **[Fig. 3]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Kuypers_etal_2018; @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015; @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975; @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink. N~2~O consumption via denitrification likely explains the N~2~O undersaturation observed at low NO~3~ concentrations, whereas denitrification was likely a source of N~2~O at higher NO~3~ concentrations.

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ [@Kuypers_etal_2018]. Nitrification may be more prevalent in the predominantly well oxygenated surface waters where the N~2~O samples were collected because denitrification is an anaeorbic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column [@Broman_etal_2021] and can be transported from aoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO~3~ relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

![Figure 3. Predicted interactive effects of surface water temperature and NO~3~ category the N~2~O saturation ratio.  A) The relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13, and B) relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category. Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

##### Temperature

The model detected a pattern of increasing N~2~O saturation with water temperature **[Fig. 3]**, which is consistent with nitrogen transformations occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ concentration. At low NO~3~ concentrations, temperature was estimated to have a relatively minor effect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature was more pronounced at greater NO~3~ concentrations, potentially reflecting the stimulatory effect of higher temperature on biological N~2~O production. The relative effect of NO~3~ was estimated to be greater than that of temperature, thus elevated water temperature may serve to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

##### Waterbody size

While NO~3~ and water temperature were associated with the N~2~O saturation ratio, N~2~O was also related to waterbody size. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreased with increasing waterbody size class **[Fig. 4]**. This pattern holds regardless of whether N~2~O was supersaturated or undersatured. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

@@

The influence of waterbody size is also suggested in the interaction between NO~3~ and surface area **[Fig. 5]**. At high NO~3~ concentrations, N~2~O is highly supersatured in small waterbodies, but moves toward equilibrium in large waterbodies **[Fig. 5A]**. At low NO~3~ concentrations, N~2~O is undersaturated in small waterbodies, but increases toward equilibrium in larger waterbodies **[Fig. 5A]**.

![Figure 5. Conditional effects plots illustrating the relationship between N~2~O saturation and A) nitrate (NO~3~ availability dependent upon lake size, and B) lake size dependent upon NO~3~ availability. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13.](manuscript_figures/sat_cond_area.tiff)

The model results suggest that NO~3~ may exert a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be associated with dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon thermal stratification strength. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] while others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

Interactions between NO~3~, water temperature, and lake size produced a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enabled identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tended to be undersaturated, whereas small waterbodies with high NO~3~ tended to be highly supersaturated **[Fig. 5]**. Large waterbodies, regardless of NO~3~ status, tended to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation may be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **[(Fig. 6A)]**, which is consistent with the relationship between N~2~O saturation and waterbody size **[(Fig. 5A,B)]**. Despite predicted widespread N~2~O undersaturation in CONUS waterbodies, the estimated national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeded the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**.

![Figure 6: A) N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. C) N~2~O flux plotted against the five lake size categories used in the survey design. Across all plots, points and bars reflect the mean (points) and 95% credible interval, respectively, of the 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Total N~2~O flux from the surface of CONUS waterbodies is a function of N~2~O emission rates and waterbody surface areas. The four smallest size classes (\< 50 Ha) in the survey design contained `r round(proportionSmallNumber, 2) * 100`% of all lakes in the population, but were estimated to have contributed only `r round((FluxBySize %>% filter(size_cat != "50_max") %>% select(estimate) %>% summarise(tot_small = sum(estimate)) %>% pull())/(FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull()) * 100, 0)`% to the N~2~O flux from CONUS waterbodies **[(Fig. 6C)]**. The reason for this pattern is two-fold. First, lakes in the smaller size classes constituted only `r round(proportionSmallSA, 2) * 100`% of total waterbody surface area, thereby limiting their N~2~O flux. Second, widespread N~2~O uptake in these size classes largely offset the very high emission rates predicted to occur in a minority of the population **[(Fig. 6A)].** Waterbodies in the largest size class (\> 50 Ha) contained only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population, but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area. This lake size class had the greatest N~2~O flux (mean = `r FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull() %>% round(0)` metric tons N~2~O year^-1^), but the flux estimate is poorly constrained with a 95% credible interval of `r FluxBySize %>% filter(size_cat == "50_max") %>% select(LCL) %>% pull()` - `r FluxBySize %>% filter(size_cat == "50_max") %>% select(UCL) %>% pull()` metric tons N~2~O year^-1^ **[(Fig. 6C)]**. While uncertainty in N~2~O emission rates was smallest for large lakes **(Fig. 6A)**, these small uncertainties propagated to large uncertainties in flux when expressed across the surface area of large waterbodies **(Fig. 6B,C)**.

@@

The mean flux estimate for CONUS waterbodies was positive (`r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^); however, the credible interval was quite broad and included zero (`r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^) so we could not conclude with statistical confidence whether CONUS waterbodies were a net source or sink of N~2~O overall. Though we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change (IPCC) estimated that CONUS drainage networks, inclusive of zero order streams through major river mouths, emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ [@epa_2023]. The IPCC estimates total N~2~O emissions from aquatic ecosystems as a fraction (emission factor) of the anthropogenic nitrogen that leaches from soils into surface water systems and does not distinguish between components of the aquatic continuum, thus a direct comparison between the IPCC estimate and the value reported here is not possible; however, the mean N~2~O  flux from lakes and reservoirs reported here comprises only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of the IPCC estimate, suggesting the IPCC estimate is too high, possibly because the IPCC model does not allow for N~2~O consumption in lakes, ponds, and reservoirs. This limitation extends to most approaches used to estimate aquatic N~2~O fluxes [@Lauerwald_etal_2023], inlcuding more recent mechanistic models [@Lauerwald_etal_2019].  Given our estimate of widespread N~2~O consumption in CONUS waterbodies, failure to include this dynamic in aquatic N~2~O models may yield biased emission estimates.

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using non-linear interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). The model form and estimated parameters suggest that NO~3~ concentration plays a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates. Our observations of dissolved N~2~O was predicted  due to high emissions from from a subset of the population.

>>>>

@@

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but recent evidence of N~2~O uptake across a broad range of waterbodies have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N2O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull()` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r propN2oSinkMean` % (95% credible interval:`r propN2oSinkLower` - `r propN2oSinkUpper`% ) of waterbodies functioning as N~2~O sinks. The model predicts dissolved N2O concentrations reasonably well based partly on interactions between nitrate (NO~3~) concentration, waterbody size, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^. Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Nitrous oxide dynamics have been investigated in streams and rivers spanning a broad range of environmental conditions, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochemistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that were at least 1 hectare in area, at least 1 meter deep, had at least 0.1 hectare of open water, and had a minimum residence time of one week[@EPAdesign] **(Fig. 1)**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **(Fig. 1)**. These data, along with the survey design information, were used to develop a Bayesian hierarchical model to predict N~2~O concentrations for all 465,897 CONUS waterbodies in the target population. At the population level, the majority of CONUS waterbodies (mean = `r propN2oSinkMean`%, 95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) were predicted to be a net sink for N~2~O. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), although the 95% credible intervals overlap across all ecoregions **(Table 1)**.

@@

The model results suggest that NO~3~ may exert a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area **(Fig. 3)**. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be associated with dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon thermal stratification strength[@Webb_etal_2019]. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] while others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

The influence of NO~3~ levels on N~2~O saturation **(Fig. 3)** may reflect the relationship between NO~3~ levels and denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ [@Kuypers_etal_2018; @Richardson_etal_2009]. The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column. Incomplete denitrification often occurs when NO~3~ is relatively abundant [@Beaulieu_etal_2015; @Beaulieu_etal_2014], likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification [@Koike_etal_1975; @Richardson_etal_2009]. When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink. N~2~O consumption via denitrification may explain the N~2~O undersaturation predicted at low NO~3~ levels, whereas denitrification may be a source of N~2~O at higher NO~3~ levels.

It is also possible that the correlation between N~2~O and NO~3~ reflects the simultaneous production of these solutes via nitrification. Nitrification, an aerobic process which can produce N~2~O as a byproduct, is the microbial oxidation of ammonium (NH~4~^+^) to NO~3~ [@Kuypers_etal_2018]. Nitrification may be more prevalent than denitrification in the predominantly well oxygenated surface waters where the N~2~O samples were collected because denitrification is an anaerobic process. However, denitrification can occur in anoxic microsites within particulate matter suspended in an aerobic water column [@Broman_etal_2021] and can be transported from aoxic sediments to the water surface via advection and diffusion [@DelSontro_etal_2018]. Thus the N~2~O - NO~3~ relationship may be reflective of some combination of N~2~O production/consumption via denitrification and N~2~O production via nitrification.

![Figure 3. Estimated interactive effects of surface water temperature (A and B) and waterbody size (B and D) with NO~3~ levels on the N~2~O saturation ratio.  A) The relationship between N~2~O saturation and surface water temperature, dependent upon nitrate (NO~3~) level.  The five, ordered, categorical levels for NO~3~ correspond to concentration ($\mu$g L^-1^) ranges in the observed data of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13. B) The relationship between N~2~O saturation and NO~3~ level conditional on surface water temperature. The NO~3~ effect was estimated at 5 quantiles of the surface temperatures observed (10%, 25%, 50%, 75%, 90%). C) The relationship between N~2~O saturation and waterbody surface area (log-transformed hectares) conditional on NO~3~ level. D) The relationship between N~2~O saturation and NO~3~ level conditional on (log) waterbody surface area (5 quantiles).  Horizontal dashed lines in each panel reflect an N~2~O saturation value of 1.](manuscript_figures/cond_effect_no3.tiff)

The estimated pattern of increasing N~2~O saturation with water temperature **(Fig. 3)** is  consistent with nitrogen transformations occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ levels. At low NO~3~ levels, temperature was estimated to have a relatively minor effect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature was more pronounced at greater NO~3~ levels, potentially reflecting the stimulatory effect of higher temperature on biological N~2~O production. The relative effect of NO~3~ was estimated to be greater than that of temperature, thus elevated water temperature may serve to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

Nitrous oxide saturation ratio was also related to waterbody size **(Fig. 3)**.  Small waterbodies with low NO~3~ levels were predicted to be undersaturated, whereas small waterbodies with high NO~3~ were predicted to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreased with increasing waterbody size class **[Fig. 4]**. This pattern holds regardless of whether N~2~O was supersaturated or undersatured. Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column [@Carpenter_1983]. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

@@

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which was expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Variability and magnitude of predicted N~2~O emission rates decreased with increasing waterbody size **(Fig. 5A)**, which is consistent with the relationship between N~2~O saturation and waterbody size **(Fig. 3C,D)**. Despite predicted widespread N~2~O undersaturation in CONUS waterbodies, the estimated national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **(Table 1)**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeded the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **(SI Fig. 1)**.

![Figure 5: A) N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 waterbodies in the target population. Across all plots, points and bars reflect the mean (points) and 95% credible interval, respectively, of the 2000 realizations of the posterior predictive distribution.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Total N~2~O flux from the surface of CONUS waterbodies is a function of N~2~O emission rates and waterbody surface areas. The four smallest size classes (\< 50 Ha) in the survey design represented `r round(proportionSmallNumber, 2) * 100`% of all lakes in the population but were estimated to have contributed only `r round((FluxBySize %>% filter(size_cat != "50_max") %>% select(estimate) %>% summarise(tot_small = sum(estimate)) %>% pull())/(FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull()) * 100, 0)`% to the N~2~O flux from CONUS waterbodies. The reason for this pattern is two-fold. First, lakes in the smaller size classes constituted only `r round(proportionSmallSA, 2) * 100`% of total waterbody surface area, thereby limiting their N~2~O flux. Second, widespread N~2~O uptake in these size classes largely offset the very high emission rates predicted to occur in a minority of the population. Waterbodies in the largest size class (\> 50 Ha) represented only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area. This lake size class was the source of `r ((FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull()) / (national_stats %>% select(post_med_fn2o) %>% pull()) * 100) %>% round(., 0)`% of the total N~2~O flux (mean = `r FluxBySize %>% filter(size_cat == "50_max") %>% select(estimate) %>% pull() %>% round(0)` metric tons N~2~O year^-1^), but the flux estimate is poorly constrained with a 95% credible interval of `r FluxBySize %>% filter(size_cat == "50_max") %>% select(LCL) %>% pull()` - `r FluxBySize %>% filter(size_cat == "50_max") %>% select(UCL) %>% pull()` metric tons N~2~O year^-1^. While uncertainty in N~2~O emission rates was smallest for large lakes **(Fig. 6A)**, these small uncertainties propagated to large uncertainties in flux when expressed across the surface area of large waterbodies **(Fig. 6B)**.

@@

The mean flux estimate for CONUS waterbodies was positive (`r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^); however, the credible interval was quite broad and included zero (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^) so we could not conclude with statistical confidence whether CONUS waterbodies were a net source or sink of N~2~O overall. Though we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change (IPCC) estimated that CONUS drainage networks, inclusive of zero order streams through major river mouths, emit `r format(ipcc.n2o.surface, scientific = FALSE, big.mark = ",")` metric tons N~2~O year^-1^ [@epa_2023]. The IPCC estimates total N~2~O emissions from aquatic ecosystems as a fraction of the anthropogenic nitrogen that leaches from soils into surface water systems but does not distinguish standing from flowing waters, thus a direct comparison between the IPCC estimate and the value reported here is not possible; however, the mean N~2~O  flux from lakes and reservoirs reported here comprises only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of the IPCC estimate, suggesting the IPCC estimate is too high, possibly because the IPCC model does not allow for N~2~O consumption in lakes, ponds, and reservoirs. This limitation extends to many mechanistic models used to estimate N~2~O fluxes from inland waters [@Lauerwald_etal_2019;@Marzadri_etal_2021;@Marzadri_etal_2017].  Given our estimate of widespread N~2~O consumption in CONUS waterbodies, failure to include this dynamic in aquatic N~2~O models may yield low-biased emission estimates [@Lauerwald_etal_2023]. Recent model parameterizations that allow for N~2~O consumption in inland waters are promising developments in this regard [@Yao_etal_2019; @Wang_etal_2023;@Maavara_etal_2019].

@@

Our research is the first to suggest that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Our observations of dissolved N~2~O were predicted reasonably well using non-linear interactions between observed water temperature, NO~3~ concentration, and waterbody size (see SI). The model form and estimated parameters suggest that NO~3~ concentration plays a role in determining whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved and equilibrium N~2~O concentrations translate to large uncertainties in flux when propagated across the surface of large waterbodies. Uncertainty in the national flux estimate was too large to determine if CONUS waterbodies functioned as a net source or sink of N~2~O. While our sample included some of the largest waterbodies in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nations largest lakes including an assessment of intra-waterbody spatial variation in N~2~O, which can be important [@Salk_etal_2019, @Borges_etal_2023], as well as careful accounting of measurement an analytical uncertainty which can propagate to large uncertainties in flux estimates.
